Scalable Property-Guided Program Regeneration

### 2025-08-01 docs/plos/draft.tex
Updates
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\title{\sys: Scalable, Property-Guided\\ Program Regeneration}
\author{Evangelos Lamprou}

@@

Three key insights shape the design of \sys.
First, modern software is highly decomposable~\cite{vfunction2024modular, isoline2018decomposition, schechter2011visualizing, breakapp:ndss:2018}, making it feasible to reason about transformations at the granularity of individual modules, libraries, or functions.
Second, the program \emph{regeneration} setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface harnesses this progress without disrupting the framework's end user.

@@

\heading{Rethinking program transformations}
% Here, talk about bidirectional analysis, how now having a program that extracts 
% a propoerty means that plugging it into sys will allow you to regenerate programs that hold or max/min-inimize over that property
Traditional transformation frameworks typically apply a one-way mapping: 
a program is analyzed to detect issues or features, and, separately, 
a hand-written rule or heuristic is applied to transform it. 
\sys enables a fundamentally different, \emph{bidirectional} view. 
Once an analysis can \emph{extract} a property,
that same analysis, when plugged into \sys's declarative query language and 
verification loop, can be used to \emph{regenerate} programs that 
satisfy, minimize, or maximize that property. 
This shifts program analyses from passive diagnostics into \emph{active drivers} 
of change: any property extractor becomes, effectively, a transformation oracle. 
The result is an extensible ecosystem where existing analyses not only tell us 
what is wrong or suboptimal, but also allow us to automatically create 
semantics-preserving, property-optimized alternatives.

@@

\heading{Availability}
The \sys early-stage prototype in available as an open-source MIT-licensed artifact at:
\begin{center}
  \url{https://github.com/blinded}
\end{center}

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
A single codebase includes many needs for transformations.
For example, a developer going through older parts of the codebase written by 
a previous team might realize that many functions are susceptible to buffer-overflow attacks.
Then, thinking about a recently incorporated JavaScript scripting component 
that allows players to make custom extensions to key parts of client-side game logic, 
suggests that perhaps code that deals with user input should all be moved inside 
the sandboxed runtime.
Furthermore, the development team would like to identify functions that could 
be broken down further to assist with performance profiling.

\heading{The state of the art and practise}
Currently, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{avgustinov2016ql, Fowler99,Mens04,Myers16} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target language.
As the programming-language landscape is becoming increasingly diverse,
and developers are building systems where multiple different languages are runtimes coexist,
these limitations cannot be ignored.
Program synthesis could be an attractive avenue that allows 
the developer to 
The ever-present limitation of traditional program synthesis techniques~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming, harp:ccs:2021, yaghmazadeh2018automated} is their 
inability to scale beyond narrowly-defined computational domains.
This scalability barrier is what statistical techniques like LLMs 
have promised to solve.
An LLM could in principle be tasked with performing all the aforementioned transformations.
However, they come with no shortage of well-studied shortcomings~(\cref{sec:intro}).
There have been systems that provide ranging amounts of control of LLM-generated 
programs~\cite{huang2023anplnaturalprogramminginteractive,li2025mccoderstreamliningmotioncontrol,barua2025pygencollaborativehumanaiapproach, tihanyi2025new},
but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strengths,
end-users to steer their output, and---more importantly---be aware when 
their outputs are \emph{wrong}.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively inside a codebase.
  In the above graph, edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied to~(\cref{sec:example}).
  Dotted edges signify omitted layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

This example is used to highlight the
diversity of transformation goals that arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

\heading{A legacy input handling function}
Zooming into the \ttt{cmd.c} file of the game codebase, 
which handles parsing and executing user commands provided by an in-game console.
This functionality is used by developers to test and debug the game, but also during regular gameplay
by game administrators or players where they can customize the game by spawning objects, changing game settings, \etc.
Such code is often notoriously difficult to write,
and there have been many instances of buffer overflow vulnerabilities being introduced because of it~\cite{CVE-2006-3400, CVE-2006-3401, CVE-2007-5248, CVE-2019-1010043}.
\begin{listing}
\begin{minted}[]{js}
#define MAX_C 1024
char *Cmd_Args(int argc, char **argv) {
 static char args[MAX_C];
 for (int i = 1; i < argc; i++) {
  strcat(args, argv[i]);
  if (i < argc - 1) strcat(args, " "); }
 return args; }
\end{minted}
\caption{A string formatting function that concatenates command-line arguments into a single string.}
\end{listing}

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               crash(p,I).
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
To the right is a program written in the \sys DSL to express this intent.
This query requires the regenerated program \ttt{P\_} to exhibit the same
input-output behavior on representative inputs that would cause a buffer overflow in the original program \ttt{P}
but not on the regenerated program \ttt{P\_}.
\sys first arrives the set of properties to extract from the
original program, given the query:
\{\prop{input(P, I)}, \prop{graph(P, I, O)}, \prop{fsignt(P, S)}\}, where
\prop{input(P, I)} means that \sys will use extract a set of inputs from the original program
by prompting an LLM with the program's source code and asking it to generate a set of inputs that
would cause a buffer overflow in the original program;
\prop{graph(P, I, O)} extracts the client-observable input-output behavior of the program
by executing the original program on the generated inputs and recording the outputs;
and \prop{fsignt(P, S)} extracts each function's signature using a language-aware parser.
After 30.4 seconds and two attempts \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\begin{wrapfigure}[5]{r}{.43\columnwidth}
\begin{minted}{prolog}
language(p_, js).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
     tr(Ic, Ijs),
     tr(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query invokes the \ttt{tr} plugin
to convert each input and output into equivalent forms suitable for the target language.
\sys first determines the minimal set of properties to extract from the
original function: \{\prop{graph(P, I, O)}\}.
It then, again, invokes \ttt{tr} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.6\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \{\prop{graph(P, I, O)}, \prop{len(P, N)}\}
from the original program
the captures captures the input-output behavior on a generated set of
inputs.
In this case, where the goal is to optimize an aspect of the program---its length---\sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes no difference after two consecutive regenerations.

@@

\heading{A benchmark suite for program regeneration}
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing---researchers and practitioners rely on ad hoc examples that are hard to compare and often
oversimplified.
This lack of a shared baseline obscures progress. 
A useful set of benchmarks should include
programs that
(1) are real and non-trivial,
(2) are written in a variety of languages, 
(3) cover a wide range of computational domains.
Each benchmark should include an original program, 
the same program that undergoes the transformation described in natural-language 
in an accompanying file, and a set of tests that verify that the transformation was successful.
Following the lead of benchmarking efforts~\cite{SPEC_CPU2017, dacapo2006, koala:atc:2025}, the suite should be made open-source, community 
first, and an established technical steering committee should guide extensions and modifications towards maintaining the suite's relevance.

@@

>>>>

\title{Scalable Property-Guided Program Regeneration}
\author{}

@@

Three key insights shape the design of \sys.
First, modern software is highly decomposable~\cite{vfunction2024modular, isoline2018decomposition, schechter2011visualizing, breakapp:ndss:2018}, making it feasible to reason about transformations at the granularity of individual modules, libraries, or functions.
Second, the program \emph{regeneration} setting provides a unique advantage: the original program serves as a concrete ground-truth, enabling semantic comparisons and verification of the transformed output, something impossible in traditional synthesis tasks.
Third, the underlying synthesis technology---particularly LLMs---is evolving rapidly; a declarative interface can harness this progress without disrupting the framework's end user.

@@

\heading{Rethinking program transformations}
% Here, talk about bidirectional analysis, how now having a program that extracts 
% a propoerty means that plugging it into sys will allow you to regenerate programs that hold or max/min-inimize over that property
Traditional transformation frameworks typically apply a one-way mapping: 
a program is analyzed to detect issues or features, and, separately, 
a hand-written rule or heuristic is applied to transform it. 
\sys enables a fundamentally different, \emph{bidirectional} view. 
Once an analysis can \emph{extract} a property,
that same analysis, when plugged into \sys's declarative query language and 
verification loop, can be used to \emph{regenerate} programs that 
satisfy, minimize, or maximize that property. 
This shifts program analyses from passive diagnostics into \emph{active drivers} 
of change: any property extractor becomes, effectively, a transformation oracle. 
This is not all, as 
its also possible to morph the extracted properties,
treating them as intermediate representations, thus further increasing 
the transformation space.
% The result is an extensible ecosystem where existing analyses not only tell us 
% what is wrong or suboptimal, but also allow us to automatically create 
% semantics-preserving, property-optimized alternatives.

@@

% \heading{Availability}
% The \sys early-stage prototype in available as an open-source MIT-licensed artifact at:
% \begin{center}
%   \url{https://github.com/blinded}
% \end{center}

@@

Consider a complex application, like the one
sketched in \cref{fig:ex-large-app}.
This single codebase encompasses different transformation needs.
For example, a developer going through older parts of the codebase 
might realize that many functions that accept user-controlled data are susceptible to buffer-overflow attacks.
Later, reflecting on the recently added JavaScript scripting component used for player-made extensions, it becomes clear
that these functions might be better moved entirely into the sandboxed runtime.
Furthermore, the development team would like to identify functions that could 
be broken down further to assist with performance profiling.

\heading{The state of the art and practise}
Right now, a developer has access to a myriad of different approaches 
to attack the above problems, but all come with shortcomings.
Code quality tools~\cite{avgustinov2016ql, Fowler99,Mens04,Myers16} and specialized transformation systems might be able to detect 
code-smells or certain classes of vulnerabilities but are often very highly 
tied to their target language.
As the programming-language landscape is becoming increasingly diverse,
and multiple different languages are runtimes coexist in a single system,
these limitations are becoming hard to ignore.
But even by detecting improvement opportunities, it still falls on the development team to implement them.
Program synthesis could be an attractive avenue that allows 
the developer to overcome the language barrier by synthesizing a new implementation that fits the refined specification.
However, the ever-present limitation of traditional program synthesis techniques~\cite{jha2010oracle, raza2018disjunctive, singh2016blinkfill,wu2023programming, harp:ccs:2021, yaghmazadeh2018automated} is their
difficulty to scale beyond narrowly-defined computational domains.
Statistical techniques like LLMs show promise in addressing this scalability barrier.
However, they come with no shortage of well-studied shortcomings~(\cref{sec:intro}).
An LLM could in principle be tasked with performing all the aforementioned transformations, and
there have been systems that provide ranging amounts of control of LLM-generated 
programs~\cite{huang2023anplnaturalprogramminginteractive,li2025mccoderstreamliningmotioncontrol,barua2025pygencollaborativehumanaiapproach, tihanyi2025new},
but there is a large gap of a unifying framework that identifies 
key parts of the problems and applies LLMs in a way that underlines their strengths,
end-users to steer their output, and---more importantly---be aware when 
their outputs are \emph{wrong}.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  % The high level of modularity present in modern applications allows \sys to be applied effectively inside a codebase.
  In the graph, edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied to~(\cref{sec:example}).
  Dotted edges signify omitted layers or dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

Consider the game application outlined in \cref{fig:ex-large-app}.
A diversity of transformation goals arise even within a single codebase:
security hardening, cross-language migration, modernization of unsafe idioms,
and structural refactoring.

\heading{A legacy input handling function}
Let's zoom into the \ttt{cmd.c} file of the game codebase,
which handles parsing and executing user commands provided by an in-game console~(\cref{lst:ex-parse}).
This functionality is used by developers to test and debug the game, but also during regular gameplay
by game administrators or players so they can customize the game by spawning objects, changing game settings, \etc
This kind of string-handling code is often notoriously difficult to write,
and there have been many instances of buffer overflow vulnerabilities being introduced because of it~\cite{CVE-2006-3400, CVE-2006-3401, CVE-2007-5248, CVE-2019-1010043}.
\begin{listing}
\begin{minted}[]{js}
#define MAX_C 1024
char *Cmd_Args(int argc, char **argv) {
 static char args[MAX_C];
 for (int i = 1; i < argc; i++) {
  strcat(args, argv[i]);
  if (i < argc - 1) strcat(args, " "); }
 return args; }
\end{minted}
\caption{A string formatting function that concatenates command-line arguments into a single string.}
  \label{lst:ex-parse}
\end{listing}

\begin{wrapfigure}[4]{r}{.53\columnwidth}
\begin{minted}{prolog}
fsignt(p_,S):-fsignt(p,S).
graph(p_,I,O):-graph(p,I,O),
               { crash(p,I); }.
:-crash(p_,I),graph(p_,I,_).
\end{minted}
\end{wrapfigure}
The developer wants to regenerate this function to eliminate this class of vulnuerabilities.
Some realizations: because the function is pure, its behavior can be fully described by input–output examples; moreover, if a buffer overflow exists, there must be inputs that trigger it.
To the right is a query written in the \sys DSL that hardens the function 
against malicious inputs.
This query requires the regenerated program \ttt{p\_} to exhibit the same
input-output behavior on both representative inputs, but have non-faulty behavior on inputs that will cause a segmentation fault to the original program \ttt{p}.
\sys first arrives at the set of properties to extract from the
original program, given the query (capital \ttt{P} here matching both the original and regenerated program):
\{\prop{graph(P, I, O)}, \prop{fsignt(P, S)}\}, where
\prop{graph(P, I, O)} extracts the client-observable input-output behavior of the program
by executing the original program on the LLM-generated inputs and recording the outputs;
\prop{crash(p, I)} signals the input generator to provide segmentation fault inducing inputs, 
and filters those that actually do;
and \prop{fsignt(P, S)} extracts each function's signature using a language-aware parser.
After 30.4 seconds and two attempts \sys synthesizes the following program:
\begin{minted}[]{js}
#define MAX_C 1024
char *Args(int argc, char **argv) {
static char args[MAX_C];
int i; size_t remaining = MAX_C - 1;
args[0] = 0;
for (i = 1; i < argc; i++) {
  strncat(args, argv[i], remaining);
  remaining = MAX_C - strlen(args) - 1;
  if (i != argc - 1 && remaining > 0) {
    strncat(args, " ", remaining);
    remaining = MAX_C - strlen(args) - 1;
  } }
  return args; }
\end{minted}

@@

\begin{wrapfigure}[5]{r}{.43\columnwidth}
\begin{minted}{prolog}
language(p_, js).
graph(p_, Ijs, Ojs) :-
     graph(p, Ic, Oc),
     tr(Ic, Ijs),
     tr(Oc, Ijs).
\end{minted}
\end{wrapfigure}
The query specifies that the regenerated program \texttt{P\_} must be
implemented in JavaScript and must preserve observable behavior over a
representative set of inputs and outputs.
To bridge differences in
representation between C and JavaScript, the query invokes the \ttt{tr} plugin
to convert each input and output into an equivalen one in the target language.
Behind-the-scenes a fine-tuned LLM will perform this translation.
\sys first determines the minimal set of properties to extract from the
original function: \{\prop{graph(P, I, O)}\}.
It then, again, invokes \ttt{tr} to produce
corresponding JavaScript representations of each example input and output, such
that the translated function will reproduce identical string formatting
behavior.
After 29.4 seconds, having generated 20 translated input-output pairs, \sys synthesizes the following JavaScript function:

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The query to the right uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a set of inputs which will be generated by an LLM instance.
\begin{wrapfigure}[2]{r}{.6\columnwidth}
\begin{minted}{prolog}
graph(p_,I,O):-graph(p,I,O).
#min N: len(p_, N).
\end{minted}
\end{wrapfigure}
\sys extracts the set of properties \{\prop{graph(P, I, O)}, \prop{len(P, N)}\}
from the original program
the captures captures the input-output behavior on a generated set of
inputs.
In this case, where the goal is to optimize an aspect of the program---its length---\sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes no difference after two consecutive regenerations.

@@

@@

\heading{A benchmark suite for program regeneration} % TODO: Fix this
For research on program regeneration to flourish, it needs a common benchmark
suite that captures the diversity and difficulty of real-world transformations.
Today, there is no equivalent of SPEC for performance or Defects4J for bug
fixing---researchers and practitioners rely on ad hoc examples that are hard to compare and often
oversimplified.
This lack of a shared baseline obscures progress. 
A useful set of benchmarks should include
programs that
(1) are real and non-trivial,
(2) are written in a variety of languages, 
(3) cover a wide range of computational domains.
Each benchmark should include an original program, 
the same program that undergoes the transformation described in natural-language 
in an accompanying file, and a set of tests that verify that the transformation was successful.
Following the lead of benchmarking efforts~\cite{SPEC_CPU2017, dacapo2006, koala:atc:2025}, the suite should be made open-source, community 
first, and an established technical steering committee should guide extensions and modifications towards maintaining the suite's relevance.
